---
title: "Spatial SARer Models, Temperature, and Sulfur"
author: "Matt Thacker"
date: "May 5, 2017"
output: html_document
---

Does reflectance due to sulfur aerosols have a relationship with local warming trends? local cooling? 

Linear and spatial SAR models are estimated between measures of local extreme temperatures and top of atmosphere radiative forcing due to sulfur aerosols. For more information on the data used, analysis of the results, and the specifications of the models estimated see the projects tab and write up also posted in this repository. 



Reading in the data and establishing near neighbor lists and a spatial weights matrix  
```{r}
library(rgdal)
library(maptools)
library(spatialreg)
library(spdep)


##define important variables

#census stations
census<- readOGR(dsn="Shapefiles/CensusStations","CensusFlux_Points")
names(census@data)[names(census@data) == "RASTERVALU"]<- "Flux"   

#k nearest neighbor list 
NLcensus<- knn2nb(knearneigh(census, k=5)) 

#spatial weights matrix
Wcensus<- nb2listw(NLcensus) 

```



Estimating naive linear models and checking for spatial autocorrelation in the error
```{r}
#Estimate OLS of relationship of interest
olsTmax<- lm(D_Tmax ~ Flux, data = census@data)
olsTmin<- lm(D_Tmin ~ Flux, data = census@data)
olsTmaxNew<- lm(TMAXNEW ~ Flux, data = census@data)
olsTminNew<- lm(TMINNEW ~ Flux, data =census@data)


#Moran's I test for Spatial Autocorrelation in relationship
miTmax<- lm.morantest(olsTmax, Wcensus, alternative = "two.sided")
miTmin<- lm.morantest(olsTmin, Wcensus, alternative = "two.sided")
miTmaxNew<- lm.morantest(olsTmaxNew, Wcensus, alternative = "two.sided")
miTminNew<- lm.morantest(olsTminNew, Wcensus, alternative = "two.sided")
miTmax
miTmin
miTmaxNew
miTminNew
```



Selecting between SARlag, SARer, or mixed model using robust Lagrange Multipier tests for spatial dependence
```{r}
#Lagrange Multiplier test for spatial lag vs. error
lmTmax <- lm.LMtests(olsTmax, Wcensus, test = "all")
lmTmin <- lm.LMtests(olsTmin, Wcensus, test = "all")
lmTmaxNew <- lm.LMtests(olsTmaxNew, Wcensus, test= "all")
lmTminNew <- lm.LMtests(olsTminNew, Wcensus, test = "all")
lmTmax
lmTmin
lmTmaxNew
lmTminNew
```



Fitting spatial SARer models
```{r}
#Estimate Spatial Simultaneous Autoregressive Error Model
EsarTmax<- errorsarlm(D_Tmax ~ Flux, data = census@data, Wcensus, method = "LU", control=list(returnHcov=FALSE))
EsarTmin<- errorsarlm(D_Tmin ~ Flux, data = census@data, Wcensus, method = "LU", control=list(returnHcov=FALSE))
EsarTmaxNew<- errorsarlm(TMAXNEW ~ Flux, data = census@data, Wcensus, method = "LU", control=list(returnHcov=FALSE))
EsarTminNew<- errorsarlm(TMINNEW ~ Flux, data = census@data, Wcensus, method = "LU", control=list(returnHcov=FALSE))
summary(EsarTmax)
summary(EsarTmin)
summary(EsarTmaxNew)
summary(EsarTminNew)
```



Examining data for trends which may contextualize or explain results.
```{r}
#attach attribute table for convience 
dat<- census@data
attach(dat)

#histograms of response variables 
hist(TMAXNEW,xlab = "Tmax_2005", main = "Tmax_2005")
hist(TMINNEW, xlab = "Tmin_2005", main = "Tmin_2005")
hist(D_Tmax, xlab = "Tmax", main = "Tmax")
hist(D_Tmin, xlab = "Tmin", main = "Tmin")

#plotted vs. Flux values
plot(Flux, D_Tmax, main = "TOA Flux vs. Tmax", ylab = "Tmax", ylim = c(0, 350), pch = 20)
plot(Flux, D_Tmin, main = "TOA Flux vs. Tmin", ylab = "Tmin", ylim = c(0,350), pch = 20)
plot(Flux, TMAXNEW, main = "TOA Flux vs. Tmax_2005", ylab = "Tmax_2005", ylim = c(0,200), pch = 20)
plot(Flux, TMINNEW, main = "TOA Flux vs. Tmin_2005", ylab = "Tmin_2005", ylim = c(0,200), pch = 20)
```

